---
title: "Vanpeppen et al. (2018) Data Analysis [PILOT]"
author: "Christine Lee"
date: "last updated at 11.20.25"
format: 
  html:
    code-fold: false
    toc: true
    toc-depth: 3
    theme: cosmo
editor: visual
---

# Introduction

\[add brief introduction of what the project will be about\]

# Setup

```{r}
# Load required packages
library(tidyverse) 
library(knitr)
library(kableExtra)

# Set theme for plots
theme_set(theme_minimal(base_size = 12))

# Set random seed for reproducibility
set.seed(123)
```

# Data Loading

Defining the pilot data files and making sure that all the data are analyzed.

```{r}
# Defining folder that contains all my pilot data
data_folder <- "pilot data"

# Obtain all the .csv files in the data_folder
all_files <- list.files(path = data_folder, pattern = "\\.csv$", full.names = TRUE)

# Read and combine all the csv files into one data set
raw_data <- all_files %>%
  map_df(~ read.csv(.x))

# Look at structure of the combined data
glimpse(raw_data)

# Look at the first few rows of the data
head(raw_data)
```

# Pre-processing Data

To ensure that the correct data is integrated,

```{r}
# Adding the data of the participants who did complete the test
data_complete <- raw_data %>%
  filter(success == TRUE)
# Filtering out participants who failed the attention check
data_attention_check <- data_complete %>%
  filter(attention_question_passed == TRUE &
           attention_effort_passed == TRUE)
```

# Scoring the Test

In this section, I am going to score the pretest & post-test for accuracy. I am working on conceptually understanding this section,

```{r}
#----------------------
# Scoring Pretest
#----------------------
# Focus on the pretest row
pretest <- data_attention_check %>%
  filter(task == "pretest")
# Inputting the correct pretest answers 
correct_Q0 <- "No correct conclusions possible"
correct_Q1 <- "blah blah"
correct_Q2 <- "blah blah"
correct_Q3 <- "blah blah"
correct_Q4 <- "blah blah"
correct_Q5 <- "blah blah"
correct_Q6 <- "blah blah"
correct_Q7 <- "blah blah"
# Score for correctedness (1 being correct, 0 being incorrect)
##Currently researching how to complete

#----------------------
# Scoring Posttest
#----------------------
##Same thing, but just for posttest variables
```

# Descriptive Stats

```{r}
# Find the mean, SD, minimum, and maximum of the pretest score
pretest_descriptives 
```
